---
title: "Pivoting in SQL"
author: "John Möller"
date: "6/11/2025"
draft: false
categories: ['SQL']
---
# Long and wide data
Consider these two tables (made up data):

```{r}
#| echo: false
library(tidyr)
library(knitr)
df <- data.frame(Store = c("Media Markt", "Media Markt",
                           "Elgiganten", "Elgiganten",
                           "Expert", "Expert"),
                 Day = c("Weekday", "Weekend",
                         "Weekday", "Weekend",
                         "Weekday", "Weekend"),
                 Status = c("Open", "Closed",
                            "Open", "Open",
                            "Closed", "Closed"))
kable(df, caption = "'Longer' table")
```

```{r}
#| echo: false
kable(tidyr::pivot_wider(df, names_from = Day, values_from = Status), caption = "'Wider' table")
```

They contain the same kind of information (except perhaps specifying the open status as specifically "Status"). Each of these forms have different advantages, thus it would be useful to be able to convert between these two formats in SQL.

# Pivot longer
```{r}
#| include: false
# Fixing SQL connection
library(DBI)
library(RSQLite)
library(here) 
db_file_path <- here("posts/PivotingSQL/database")
con <- dbConnect(RSQLite::SQLite(), dbname = db_file_path)
```


Consider the table birthstatistics which first lines looks like this:

```{sql connection=con}
SELECT * FROM birthstatistics
LIMIT 5
```

Let's say we want to do a bunch of operations where we specify mathematical
constraints on the year number. Let's say one of them is to get the data
where the year is an even number.
If there was a column called Year we could specify "WHERE Year % 2 = 0".
But we can't do that as each year is a column and not a cell value in a column.
 Thus we would like instead a form where there was a column called "Year".
Then we can have the values in each cell be inside a single column that we will call "Persons". This is one way to achieve that in SQL (not the prettiest way, but one of the most straightforward):
```{sql connection=con}
CREATE View long_table AS
SELECT region, sex, 'foreign/Swedish', 2002 as Year, [2002] as Persons FROM birthstatistics
UNION ALL
SELECT region, sex, 'foreign/Swedish', 2003 as Year, [2003] as Persons FROM birthstatistics
UNION ALL
SELECT region, sex, 'foreign/Swedish', 2004 as Year, [2004] as Persons FROM birthstatistics
UNION ALL
SELECT region, sex, 'foreign/Swedish', 2005 as Year, [2005] as Persons FROM birthstatistics
UNION ALL
SELECT region, sex, 'foreign/Swedish', 2006 as Year, [2006] as Persons FROM birthstatistics
UNION ALL
SELECT region, sex, 'foreign/Swedish', 2007 as Year, [2007] as Persons FROM birthstatistics
UNION ALL
SELECT region, sex, 'foreign/Swedish', 2008 as Year, [2008] as Persons FROM birthstatistics
UNION ALL
SELECT region, sex, 'foreign/Swedish', 2009 as Year, [2009] as Persons FROM birthstatistics
UNION ALL
SELECT region, sex, 'foreign/Swedish', 2010 as Year, [2010] as Persons FROM birthstatistics
UNION ALL
SELECT region, sex, 'foreign/Swedish', 2011 as Year, [2011] as Persons FROM birthstatistics
UNION ALL
SELECT region, sex, 'foreign/Swedish', 2012 as Year, [2012] as Persons FROM birthstatistics
UNION ALL
SELECT region, sex, 'foreign/Swedish', 2013 as Year, [2013] as Persons FROM birthstatistics
UNION ALL
SELECT region, sex, 'foreign/Swedish', 2014 as Year, [2014] as Persons FROM birthstatistics
UNION ALL
SELECT region, sex, 'foreign/Swedish', 2015 as Year, [2015] as Persons FROM birthstatistics
UNION ALL
SELECT region, sex, 'foreign/Swedish', 2016 as Year, [2016] as Persons FROM birthstatistics
UNION ALL
SELECT region, sex, 'foreign/Swedish', 2017 as Year, [2017] as Persons FROM birthstatistics
UNION ALL
SELECT region, sex, 'foreign/Swedish', 2018 as Year, [2018] as Persons FROM birthstatistics
UNION ALL
SELECT region, sex, 'foreign/Swedish', 2019 as Year, [2019] as Persons FROM birthstatistics
UNION ALL
SELECT region, sex, 'foreign/Swedish', 2020 as Year, [2020] as Persons FROM birthstatistics;
```

Now we can do a statement like the following:
```{sql connection=con}
SELECT * FROM long_table
WHERE Year % 2 = 0 AND sex = "girls" AND Persons > 100000
LIMIT 5;
```

# Pivot wider
Now let's say we want boys and girls to be columns. We can achieve that with the following:

```{sql connection=con}
CREATE VIEW My_view AS
SELECT region, 'foreign/Swedish', Year,
  SUM(CASE WHEN sex = 'boys' THEN Persons ELSE 0 END) AS Boys,
  SUM(CASE WHEN sex = 'girls' THEN Persons ELSE 0 END) AS Girls
FROM long_table
GROUP BY region, 'foreign/Swedish', Year
```

Why are we using SUM? Well in this example we kind of assume that given a distinct
combination of the columns 'region', 'foreign/Swedish' and 'Year', then there only exists on row where 'sex' is specified as for example boy, and the value in 'Persons' is the value we'll put under the column 'boy' in the widened table with that combination. But what if there are multiple? One way for example to deal with it is to just pick the first row and ignore the others. But one way that usually generalizes well is to sum all of them (perhaps each row symbolizes a sub-region for example).

Here's an example of a query using the the view with widened data:
```{sql connection=con}
SELECT * FROM My_View
WHERE Year % 2 = 0
LIMIT 5
```

As an example let's create two graphs for the total boys through the years and the total
girls through the years.

```{sql connection=con}
CREATE VIEW My_total AS
SELECT Year, SUM(Boys) AS Boys, SUM(Girls) AS Girls
FROM My_view
GROUP BY Year
```


```{r}
library(ggplot2)
my_view_df <- dbGetQuery(con, "SELECT * FROM My_total")
ggplot(my_view_df, aes(Year, Boys)) + geom_line()
ggplot(my_view_df, aes(Year, Girls)) + geom_line()
```


```{sql connection=con}
#| include: false
DROP VIEW long_table;
```

```{sql connection=con}
#| include: false
DROP VIEW My_view;
```
```{sql connection=con}
#| include: false
DROP VIEW My_total
```

```{r}
#| include: false
dbDisconnect(con)
```

